pirouette: Determining the error BEAST2 makes in inferring a phylogeny

### 2019-03-07 article.tex
Use 'ht' float placement, as indicated by Overleaf
Added 4th example, intermediate between 3rd and 5th

\begin{figure}[h]
  \includegraphics[width=\textwidth]{figure_bd.png}
  \caption{The Yule tree, as created by listing \ref{lst:create_yule_tree}}
\end{figure}

@@

\begin{figure}[h]
  \includegraphics[width=\textwidth]{figure_example_1.png}
  \caption{
    Example 1: the error BEAST2 makes from a phylogeny 
    when the generative and inference model are the same.
    The x-axis specifies the tree type (which can be 'true' or 'twin'), 
    the y-axis shows the error. 
    The boxplot shows the minimum, first quartile, median, third 
    quartile and maximum error, of which the values are displayed 
    in the adjacent label
  }
  \label{fig:example_1}
\end{figure}

@@

\begin{lstlisting}[
  language=R, 
  floatplacement=ht, 
  frame=single, 
  label = {lst:unknown_phylogeny},
  caption = A phylogeny generated by an unknown diversification model
]
phylogeny  <- ape::read.tree(
  text = "(((A:8, B:8):1, C:9):1, ((D:8, E:8):1, F:9):1);"
)
\end{lstlisting}

@@

\begin{lstlisting}[language=R, floatplacement=ht, frame=single]
experiments <- create_all_experiments()
\end{lstlisting}

@@

\begin{figure}[h]
  \includegraphics[width=\textwidth]{figure_example_2.png}
  \caption{
    Example 2: the error BEAST2 makes from a phylogeny when
    picking the best inference model.
    The x-axis specifies the tree type (which can be 'true' or 'twin'), 
    the y-axis shows the error
  }
  \label{fig:example_2}
\end{figure}

@@

\begin{lstlisting}[language=R, floatplacement=ht, frame=single]
twinning_params <- create_twinning_params()
\end{lstlisting}

@@

\begin{figure}[h]
  \includegraphics[width=\textwidth]{figure_example_3.png}
  \caption{
    Example 3: the error made by BEAST2 from a phylogeny, picking the best inference model versus the control.
    Here, the twin column shows the error BEAST2 makes starting from the original tree versus the twin. 
    The x-axis specifies the tree type (which can be 'true' or 'twin'), the y-axis shows the error.
    The boxplot shows the minimum, first quartile, median, third 
    quartile and maximum error, of which the values are displayed 
    in the adjacent label
  }
  \label{fig:example_3}
\end{figure}

@@

"What is the error made by BEAST2 from a phylogeny, 
when the generative model is known, 
compared to candidates with similar inference models, 
in relationship to the background error?"
\richel{
  I think this is a nonsense research question and example:
  we should use a tree of unknown speciation model instead.
  Only then does twinning make sense
}

Here, same as example 2, we start with a tree generated by the Yule process, 
by using the code shown in listing \ref{lst:create_yule_tree}.
The first step in \verb;pirouette; is to simulate a DNA alignment 
from the given phylogeny, for which we will use the same code 
as shown in listing \ref{lst:create_alignment}.

\begin{table}
  \begin{tabular}{ | c | c | c | l | }
    \hline
    \textbf{model type} & \textbf{run if} & \textbf{measure evidence} & \textbf{inference model} \\ 
    \hline
    generative & always         & yes & JC, strict, Yule \\
    candidate  & best candidate & yes & JC, strict, BD \\
    candidate  & best candidate & yes & JC, strict, CCP \\
    candidate  & best candidate & yes & JC, strict, CEP \\
    \hline
  \end{tabular}
  \caption{
    Experimental setup to answer the fourth research question.
    JC: Jukes-Cantor site model.
    strict: strict clock model.
    Yule: Yule (pure-birth) tree prior.
    BD: birth-death tree prior.
    CCP: coalescent constant-population tree prior.
    CEP: coalescent exponential-population tree prior.
  }
  \label{tab:experiment_4}
\end{table}

\begin{lstlisting}[language=R, floatplacement=ht, frame=single]
generative_experiment <- create_experiment(
  model_type = "generative",
  run_if = "always",
  do_measure_evidence = TRUE
)
\end{lstlisting}

\begin{lstlisting}[language=R, floatplacement=ht, frame=single]
candidate_experiments <- create_all_experiments(
  site_models = list(create_jc69_site_model()),
  clock_models = list(create_strict_clock_model()),
  tree_priors = list(
    create_bd_tree_prior(), 
    create_ccp_tree_prior(), 
    create_cep_tree_prior()
  )
)
\end{lstlisting}

\begin{lstlisting}[language=R, floatplacement=ht, frame=single]
experiments <- c(list(generative_experiment), candidate_experiments)
\end{lstlisting}

\begin{lstlisting}[language=R, floatplacement=ht, frame=single]
pir_params <- create_pir_params(
  alignment_params = alignment_params,
  experiments = experiments,
  twinning_params = create_twinning_params()
)
\end{lstlisting}

@@

\begin{figure}[h]
  \includegraphics[width=\textwidth]{figure_example_4.png}
  \caption{
    Example 4: the error made by BEAST2 from a phylogeny, when the generative model is known, compared to candidates with similar inference models, in relationship to the background error.
    The x-axis specifies the tree type, the y-axis shows the error.
    The boxplot shows the minimum, first quartile, median, third 
    quartile and maximum error, of which the values are displayed 
    in the adjacent label
  }
  \label{fig:example_4}
\end{figure}

>>>>

\begin{figure}[ht]
  \includegraphics[width=\textwidth]{figure_bd.png}
  \caption{The Yule tree, as created by listing \ref{lst:create_yule_tree}}
\end{figure}

@@

\begin{figure}[ht]
  \includegraphics[width=\textwidth]{figure_example_1.png}
  \caption{
    Example 1: the error BEAST2 makes from a phylogeny 
    when the generative and inference model are the same.
    The x-axis specifies the tree type (which can be 'true' or 'twin'), 
    the y-axis shows the error. 
    The boxplot shows the minimum, first quartile, median, third 
    quartile and maximum error, of which the values are displayed 
    in the adjacent label
  }
  \label{fig:example_1}
\end{figure}

@@

\begin{lstlisting}[
  language=R, 
  floatplacement=ht, 
  frame=single, 
  label = {lst:unknown_phylogeny},
  caption = A phylogeny generated by an unknown diversification model
]
phylogeny  <- ape::read.tree(
  text = "(((A:8, B:8):1, C:9):1, ((D:8, E:8):1, F:9):1);"
)
\end{lstlisting}
\begin{figure}[ht]
  \includegraphics[width=\textwidth]{tree_unknown.png}
  \caption{The tree derived from an unknown diversification process, 
    as created by listing \ref{lst:unknown_phylogeny}
  }
\end{figure}

@@

\begin{lstlisting}[
  language=R, 
  floatplacement=ht, 
  frame=single, 
  label = {lst:create_all_experiments},
  caption = Create all 40 candidate experiments
]
experiments <- create_all_experiments()
\end{lstlisting}

@@

\begin{figure}[ht]
  \includegraphics[width=\textwidth]{figure_example_2.png}
  \caption{
    Example 2: the error BEAST2 makes from a phylogeny when
    picking the best inference model.
    The x-axis specifies the tree type (which can be 'true' or 'twin'), 
    the y-axis shows the error
  }
  \label{fig:example_2}
\end{figure}

@@

\begin{lstlisting}[
  language=R, 
  floatplacement=ht, 
  frame=single,
  label = {lst:create_twinning_params},
  caption = Create the default twinning parameters
]
twinning_params <- create_twinning_params()
\end{lstlisting}

@@

\begin{figure}[ht]
  \includegraphics[width=\textwidth]{figure_example_3.png}
  \caption{
    Example 3: the error made by BEAST2 from a phylogeny, picking the best inference model versus the control.
    Here, the twin column shows the error BEAST2 makes starting from the original tree versus the twin. 
    The x-axis specifies the tree type (which can be 'true' or 'twin'), the y-axis shows the error.
    The boxplot shows the minimum, first quartile, median, third 
    quartile and maximum error, of which the values are displayed 
    in the adjacent label
  }
  \label{fig:example_3}
\end{figure}

@@

"What is the error made by BEAST2 from a phylogeny 
derived from an unknown speciation model,
when selecting the best inference model, 
in relationship to the background error?"
\richel{
  I inserted this example research question, 
  as think this is natural intermediate between the 
  previous and next/final example
}

Here, same as example 2 and 3, we start with the phylogeny 
from an unknown speciation model (listing \ref{lst:unknown_phylogeny}), 
The first step in \verb;pirouette; is to simulate a DNA alignment 
from the given phylogeny, for which we will use the same code 
as shown in listing \ref{lst:create_alignment}.

In the second step we specify we want all 
40 inference models to compete, by 
calling \verb;create_all_experiments;,
as done in listing \ref{lst:create_all_experiments}.

We choose again to use sensible defaults for the third (the BEAST2 inference) 
and fourth (measuring the error) steps, 
as we previously did to answer the second research question.

To get an idea of a baseline error made by BEAST2, 
we also make use of the twinning option by using the
default created twinning parameters, as shown in 
listing \ref{lst:create_twinning_params}.

@@

\begin{figure}[ht]
  \includegraphics[width=\textwidth]{figure_example_4.png}
  \caption{
    Example 4: the error made by BEAST2 from a phylogeny, when the generative model is known, compared to candidates with similar inference models, in relationship to the background error.
    The x-axis specifies the tree type, the y-axis shows the error.
    The boxplot shows the minimum, first quartile, median, third 
    quartile and maximum error, of which the values are displayed 
    in the adjacent label
  }
  \label{fig:example_4}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Usage: Example Research Question 5}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

A fifth research question that \verb;pirouette; can answer, is:

"What is the error made by BEAST2 from a phylogeny, 
when the generative model is known, 
compared to candidates with similar inference models, 
in relationship to the background error?"

In this example, we use a tree generated by the Yule process.
Instead of using the common Yule tree as in example 2, we'll use
a Yule tree of low (actually: zero) likelihood: 
the one as used in examples 3 and 4.
by using the code shown in listing \ref{lst:create_yule_tree}.
The first step in \verb;pirouette; is to simulate a DNA alignment 
from the given phylogeny, for which we will use the same code 
as shown in listing \ref{lst:create_alignment}.

\begin{table}
  \begin{tabular}{ | c | c | c | l | }
    \hline
    \textbf{model type} & \textbf{run if} & \textbf{measure evidence} & \textbf{inference model} \\ 
    \hline
    generative & always         & yes & JC, strict, Yule \\
    candidate  & best candidate & yes & JC, strict, BD \\
    candidate  & best candidate & yes & JC, strict, CCP \\
    candidate  & best candidate & yes & JC, strict, CEP \\
    \hline
  \end{tabular}
  \caption{
    Experimental setup to answer the fourth research question.
    JC: Jukes-Cantor site model.
    strict: strict clock model.
    Yule: Yule (pure-birth) tree prior.
    BD: birth-death tree prior.
    CCP: coalescent constant-population tree prior.
    CEP: coalescent exponential-population tree prior.
  }
  \label{tab:experiment_5}
\end{table}

\begin{lstlisting}[language=R, floatplacement=H, frame=single]
generative_experiment <- create_experiment(
  model_type = "generative",
  run_if = "always",
  do_measure_evidence = TRUE
)
\end{lstlisting}

\begin{lstlisting}[language=R, floatplacement=H, frame=single]
candidate_experiments <- create_all_experiments(
  site_models = list(create_jc69_site_model()),
  clock_models = list(create_strict_clock_model()),
  tree_priors = list(
    create_bd_tree_prior(), 
    create_ccp_tree_prior(), 
    create_cep_tree_prior()
  )
)
\end{lstlisting}

\begin{lstlisting}[language=R, floatplacement=H, frame=single]
experiments <- c(list(generative_experiment), candidate_experiments)
\end{lstlisting}

\begin{lstlisting}[language=R, floatplacement=H, frame=single]
pir_params <- create_pir_params(
  alignment_params = alignment_params,
  experiments = experiments,
  twinning_params = create_twinning_params()
)
\end{lstlisting}

\begin{lstlisting}[language=R, floatplacement=H, frame=single]
errors <- pir_run(
  phylogeny,
  pir_params = pir_params
)
\end{lstlisting}

\begin{lstlisting}[language=R, floatplacement=H, frame=single]
pir_plot(errors)
\end{lstlisting}

The resulting figure is shown in figure \ref{fig:example_5}.

\begin{figure}[h]
  \includegraphics[width=\textwidth]{figure_example_5.png}
  \caption{
    Example 5: the error made by BEAST2 from a phylogeny, when the generative model is known, compared to candidates with similar inference models, in relationship to the background error.
    The x-axis specifies the tree type, the y-axis shows the error.
    The boxplot shows the minimum, first quartile, median, third 
    quartile and maximum error, of which the values are displayed 
    in the adjacent label
  }
  \label{fig:example_5}
\end{figure}
